---
title: "Tutorial for R Package TBcompass"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 4
  word_document:
    toc: true
    toc_depth: '4'
vignette: >
  %\VignetteIndexEntry{TBcompass_vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial introduces the TBcompass package and walks through an example of using TBcompass to classify tuberculosis (TB) disease severity. 

First we install TBcompass with:
``` r
# install.packages("devtools")
devtools::install_github("samalatesta/TBcompass")
```

If we want the results to match this tutorial, we first set the seed to match: 
```{r seed}
set.seed(0)
```


We then load the package by running:
```{r setup}
library(TBcompass)
```

We often collect information from multiple data sources such as symptoms, sputum culture, sputum smear microscopy, and chest radiography during the process of diagnosing patients with TB. While this information is used to inform TB diagnosis, it is also strongly correlated with TB disease severity. Common approaches to categorizing TB severity include using smear microscopy or cavitation on chest x-ray alone offering only a binary view of the TB disease spectrum. We aimed to develop a method that integrates all available data to classify disease severity and provide a more detailed understanding TB disease severity at clinical presentation. 

Data-driven disease progression modeling offers methods to reconstruct a disease timeline and estimate the trajectory along this timeline informed directly by observed data (Oxtoby, 2023). There have been numerous adaptations and extensions of the event-based model (EBM) for disease progression that have primarily been developed for the purposes of understanding the progression and trajectory of different neurodegenerative diseases (Fonteijn et al., 2012; Venkatraghavan et al., 2019; Young et al., 2014). 

We leverage the EBM framework to classify different TB disease severity states at time of TB diagnosis. A benefit of the EBM is we only require cross-sectional data to fit the model which aligns with observing individuals only at time of TB diagnosis. The EBM allows us to integrate all clinical information to obtain a predicted disease stage that we use as a proxy for TB severity. We present the TBcompass package to implement our method. We provide this tutorial as a guide for conducting data analysis with TBcompass. 

## Analysis Workflow
We recommend the following workflow as a general guide for analyzing data with TBcompass. Functions required are listed in parentheses:      
1. Visualize clinical measures (plot_events())      
2. Estimate model (fit_ebm())     
3. Check likelihood ascent plot (plot_likes())     
4. Reestimate model for m bootstrap resamples (bootstrap_seq())     
5. Visualize uncertainty by making positional variance diagram (pvd())     
6. Obtain predicted disease stage for each observation and plot distribution (get_stage() and plot_stage())     

## Example Data Set   
We provide a synthetic data set TBcompassData in the TBcompass package. We will use this data set to walk through an example of analyzing data with TBcompass to classify TB disease severity.     

### Description 
In the table below, we provide a description of each variable included in the data set. 

Variable | Description |
:-------:| :--------------: | 
Index | Unique row identifier | 
stage | True disease stage | 
X1 | Indicator for event 1 in clinical measure 1 | 
X2 | Indicator for event 2 in clinical measure 1 |
X3 |  Indicator for event 1 in clinical measure 2 |
X4 |  Indicator for event 2 in clinical measure 2 | 
X5 |  Indicator for event 3 in clinical measure 2 | 
X6 |  Indicator for event 1 in clinical measure 3 |
C1 |  Binary covariate |
C2 |  Continuous covariate |  <br> 


The data set contains 250 rows where each row represents data observed on a unique individual. There are 10 columns in the data set. For each individual, we record the value for 6 disease events corresponding to 3 clinical measures, and we include 2 covariates. The motivation for including covariates is we are often interested in associations between different variables with disease severity. Each covariate was simulated to be positively correlated with disease stage. 

```{r TBcompassData str}
data(TBcompassData)
str(TBcompassData)
```

### True event sequence
Here is the ground truth sequence of events that was used to generate TBcompassData. There are 4 disease stages. Events X6, X1, and X3 occur simultaneously at stage 1: 

Event | Stage |
:-------:| :--------------: | 
X1 | 1 | 
X2 | 2 |
X3 | 1|
X4 | 3 | 
X5 | 4 | 
X6 | 1 | <br> 


**Important Note**      
  In theory we can include as many clinical measures and events as possible when fitting the EBM. The performance of the method decreases with smaller sample sizes, a larger number of events, and lower values of p for each event. We should consider these factors when choosing data to estimate the EBM. An additional challenge includes non-convergence when estimating the maximum likelihood sequence. This can potentially be mitigated by increasing the number of iterations during estimation but comes at the cost of increased computation time. 


### Data Summary 


Below we provide the first 6 rows of our example data set. The first observation in this data set is at disease stage 2. Events X1, X2, X3, and X6 have occurred as indicated with the value 1, and events X4 and X6 have not occurred which is indicated by the value 0. We generated this data set setting p to 0.9 so each individual’s data may not align perfectly with their true disease stage. For example, observation 3 and 4 are both at stage 1. Given the ground truth sequence, we would expect X1, X3, and X6 to all have the value 1. For observation 4 this is true, but for observation 3, X6 is equal to 0. 
```{r TBcompassData first rows}
head(TBcompassData)
```

Here is the distribution of disease stage. We can summarize true disease stage because we are using a simulated data set. We simulated disease stage originally using a uniform distribution, so we expect all stages to be approximately equally observed in the data. Notice here we have a disease stage 0 whereas in our true event sequence above we did not. This is because stage 0 is equivalent to no events having occurred. 
```{r stage dist}
prop.table(table(TBcompassData$stage))
```

Here are the distributions for each covariate.
```{r data dist}
prop.table(table(TBcompassData$C1))
fivenum(TBcompassData$C2)
```
## Example analysis

### Visualize clinical events
Prior to fitting the model, it is good practice to first visualize the data we are using. We can plot our data using the plot_events function. This function first calculates the frequency of each event occurring and orders the events from most frequent to least frequent. The ordered events are the expected disease sequence for the data set as a greater event frequency should correspond to earlier disease stages. We then sort observations by the ordered events. We then plot each observation to visualize how well the data follow the expected disease sequence. The x-axis corresponds to observations after sorting and the y-axis corresponds to clinical events ordered by frequency. Blue indicates the event has occurred and white indicates the event has not occurred. If our data aligns well with the expected disease sequence, we expect the top left area of the plot to be mostly white and the bottom right to be consistently blue.  
  
We present the plot from  running the plot_events function with TBcompassData. The y axis is ordered as X1, X6, X3, X2, X4, X5. This aligns with our ground truth sequence where X1, X6, and X3 occur at stage 1 followed by X2, X4, and X5 at stages 2, 3, and 4, respectively. We see our data align well with the ground truth sequence overall with few events occurring in the upper left hand portion of the plot. This is a result of us using a value of .9 for all clinical events when we simulated our data set. If p was lower, we would expect to see more variation. 
```{r plot events}
plot_events(TBcompassData, id.var="Index", event.vars=c("X1", "X2", "X3", "X4", "X5", "X6"))
```

### Estimate model
We fit the EBM using the fit_ebm function. Prior to running the model, the data set should be formatted such that each event has a  recorded value of 1 to indicate the event has occurred or 0 otherwise. If there are any observations with missing data they are excluded prior to model fitting. We require the user to also input a data frame that provides the event name for each clinical measure and event that corresponds to a column name in the data set used to fit the model.We provide the information data frame for TBcompassData that will use to fit the model as an example. We must also specify a value of p for each event.
```{r measures}
measure_info <- data.frame(clinical_measure=c("measure1", "measure1", "measure2", "measure2", "measure2", "measure3"),event_number = c(1,2,1,2,3,1), event_name=c(c("X1", "X2", "X3", "X4", "X5", "X6")))

print(measure_info)
```
The fit_ebm function returns a list. One object in the list called ml_seq is a data frame with 4 columns (measure, event_number, event_name, est_seq). The stage column in this data frame is the maximum likelihood sequence. We present the maximum likelihood sequence for TBcompassData. We ran the model with only 3 start points and 250 iterations but in practice the minimum number of start points should be 10 with at least 2000 iterations. From this output, the maximum likelihood sequence is ({X6, X1, X3}, {X2}, {X4},{X5}) which is the same as our ground truth sequence. 
```{r fit ebm}
model1 = fit_ebm(data=TBcompassData[,c(3:8)], p_vec=c(.9,.9,.9,.9,.9,.9), clinical_info=measure_info, nstart=3,initial_iter= 250)

#print maximum likelihood estimate
print(model1$ml_seq)
```

   
### Likelihood ascent
After estimating the model, we assess model convergence by plotting the likelihood ascent for each start point using the plot_likes function. The log-likelihoods for all iterations can be accessed from the saved model object. For example, we saved our model as model1 above and we can call model1$loglikes which is a dataframe that contains all log-likelihoods across all iterations and start points. We present the likelihood ascent plot for estimating the model with TBcompassData. We see all start points converged within the first 50 iterations. When running the EBM with real data, it is likely the EBM will not converge until at least 1000 iterations. When fitting the model we recommend checking convergence by plotting the likelihood ascent and if convergence is very slow or does not occur then to increase the total number of iterations and/or start points and rerun the model. 

```{r likelihood ascent}
#first 6 rows of loglikes data frame
head(model1$loglikes)

#plot log-likelihoods for all iterations and start points
plot_likes(model1$loglikes)
```

### Bootstrap resamples
After fitting the model, we recommend constructing a positional variance diagram to visualize uncertainty in the maximum likelihood sequence. We provide the function bootstrap_seq to refit the model on a chosen number of resamples with replacement from the observed data. We recommend using at least 100 resamples. This function requires the same inputs as fit_ebm with the additional argument z which is the total number of bootstrap resamples. The function returns a data frame where each row represents the sequence estimated for 1 resample from our observed data. We ran bootstrap_seq with TBcompassData and our estimated sequence above. We present the first 6 rows of output. If our model fits the data well, we expect a majority of the bootstrap sequences to be the same as our maximum likelihood sequence. Like with fit_ebm, in practice a mininum of 1000 iterations and 10 start points should be used. 
```{r bootstrap sequences}

boot_seqs = bootstrap_seq(z=25,data=TBcompassData[,c(3:8)], p_vec=c(.9,.9,.9,.9,.9,.9), clinical_info=measure_info, nstart=1,initial_iter= 250)

head(boot_seqs)
```

### Positional variance diagram
After obtaining bootstrap sequences, we plot them in a positional variance diagram (PVD) to visualize the location and degree of uncertainty for our maximum likelihood sequence using the function pvd. The function requires a data frame of bootstrap sequences and the maximum likelihood sequence as input.  The function returns a heatmap where the x-axis is disease stage and the y axis is each event in order of the maximum likelihood sequence. Each square in the plot represents the proportion of bootstrap sequences where each event occurs at each disease stage. Darker colors indicate a higher proportion/degree of certainty. The disease stage estimated for each event in the maximum likelihood sequence is outlined in black. We expect the highest proportion of bootstrap samples for each event to occur at the stage where the event occurs in the maximum likelihood sequence.    
   
We provide the PVD for the bootstrap samples for TBcompassData below. A majority of our bootstrap sequences were equivalent to the maximum likelihood sequence. We observe uncertainty for events X3 to X5 where a small proportion of sequences estimated these events to occur one stage later than the maximum likelihood sequence. It is important to note that the x axis for the pvd ranges from 1 to 6 to account for the fact that some bootstrap sequences may have a different number of stages than our maximum likelihood estimate. The maximum number of stages is always equal to the total number of events; for our example data set this is 6. 
```{r positional variance diagram}
pvd(boot_seqs = boot_seqs, ml=model1$ml_seq)
```

### Predicted disease stage
Our primary goal with using the EBM to understand TB disease severity is to obtain a predicted disease stage for each individual in our sample. We include the function get_stage which calculates the stage with greatest probability given an individual’s observed data and the estimated maximum likelihood sequence. Additionally, we include the function plot_stage that generates a bar plot for the distribution of predicted disease stage. We present the plot for predicted disease stage for our example data set. We see the disease stage distribution is approximately uniform which is expected since we used a uniform distribution to generate the true disease stage. In practice, if we are using data from TB clinics, we would expect to see a left-skewed distribution where we disproportionately observe individuals with more severe disease. 
 
     
Predicted disease can be analyzed as an outcome as researchers are frequently interested in factors associated with increased disease severity. Disease stage can also be use in secondary analysis as a predictor when modeling other outcomes of interest. 
```{r disease stage}
#save estimated sequence from model
est_seq=model1$ml_seq

#calculate predicted disease stage
stagedf <- get_stage(data=TBcompassData[,c(3:8)], est_seq, c(.9,.9,.9,.9,.9,.9))

#print first rows of data frame that includes predicted disease stage
head(stagedf)

#plot predicted disease stage
plot_stage(pred_stage=stagedf$pred_stage)
```


## Conclusion     
In this tutorial, we illustrated how to use the TBcompass package to analyze data to classify TB severity using clinical measures observed at time of TB diagnosis. With our example data set TBcompassData, we demonstrated how to do the following:   
  
* Visualize clinical events   
* Estimate the disease event sequence   
* Check the likelihood ascent plot for model convergence  
* Generate bootstrap resamples and plot the positional variance diagram to visualize uncertainty  
* Calculate predicted disease stage and plot the distribution
     
       
We conclude with some considerations and best practices for practitioners wanting to analyze data with TBcompass. 

* The method we present was motivated by prior work in event-based modeling that was largely used to understand the progression of neurodegenerative diseases. When applying our method to TB data, we are not concerned with drawing conclusions about disease progression. We are using the EBM to classify disease severity at time of diagnosis only. It would not be appropriate to comment on disease progression as we know many assumptions for the EBM are not met with TB. 
* Fitting the EBM requires collecting data that is strongly correlated with disease severity. For TB disease, we include measures that are well-established as predictors of treatment outcome and TB sterilization (cavitation, smear status, etc) and that are often used individually to stratify patients into categories that represent advanced versus early disease.   
* Fitting the EBM also requires the user to input a value for p for each event included in the model. With TB diagnostic measures and chest x-ray data, we expect p to be close to 1. The user should always run a sensitivity analysis where they decrease p from their original model fit to assess the sensitivity of the maximum likelihood sequence to p.   
* In theory we can include as many clinical measures and events as possible when fitting the EBM. The performance of the method decreases with smaller sample sizes, a larger number of events, and lower values of p for each event. We should consider these factors when choosing data to estimate the EBM. An additional challenge includes non-convergence when estimating the maximum likelihood sequence. This can potentially be mitigated by increasing the number of iterations during estimation but comes at the cost of increased computation time. The user should always check the likelihood ascent plot to determine if convergence was reached with the chosen number of iterations.   
* Finally, it is always best to visualize the observed data before fitting the model which also provides the user with an expected disease sequence to compare to the maximum likelihood estimate. This is especially helpful if we are using the EBM as a hypothesis generating tool and do not already have a strong understanding about disease sequence or severity.  
* It is important to note that we present TBcompass and this tutorial in the context of classifying TB disease severity but the the model and analysis we present are not specific to TB. The methods we developed could be used to understand disease dynamics for other chronic diseases that progress over time where we are able to measure events that are strongly correlated with the disease process. 


