---
title: "Tutorial for R Package TBebm"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 4
  word_document:
    toc: true
    toc_depth: '4'
vignette: >
  %\VignetteIndexEntry{Tbebm_vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial introduces the TBebm package and walks through an example of using TBebm to classify tuberculosis (TB) disease severity. 

First we install TBebm with:
``` r
# install.packages("devtools")
devtools::install_github("samalatesta/TBebm")
```

If we want the results to match this tutorial, we first set the seed to match: 
```{r seed}
set.seed(0)
```


We then load the package by running:
```{r setup}
library(TBebm)
```

## Analysis Workflow
We recommend the following workflow as a general guide for analyzing data with TBebm. Functions required are listed in parentheses:      
1. Visualize clinical measures (plot_events())      
2. Estimate model (fit_ebm())     
3. Check likelihood ascent plot (plot_likes())     
4. Reestimate model for m bootstrap resamples (bootstrap_seq())     
5. Visualize uncertainty by making positional variance diagram (pvd())     
6. Obtain predicted disease stage for each observation and plot distribution (get_stage() and plot_stage())     

## Example Data Set   
We provide a synthetic data set TBebmData in the TBebm package. We will use this data set to walk through an example of analyzing data with TBebm to classify TB disease severity.     

### Description 
In the table below, we provide a description of each variable included in the data set. 

Variable | Description |
:-------:| :--------------: | 
Index | Unique row identifier | 
stage | True disease stage | 
X1 | Indicator for event 1 in clinical measure 1 | 
X2 | Indicator for event 2 in clinical measure 1 |
X3 |  Indicator for event 1 in clinical measure 2 |
X4 |  Indicator for event 2 in clinical measure 2 | 
X5 |  Indicator for event 3 in clinical measure 2 | 
X6 |  Indicator for event 1 in clinical measure 3 |
C1 |  Binary covariate |
C2 |  Continuous covariate |  <br> 


The data set contains 250 rows where each row represents data observed on a unique individual. There are 10 columns in the data set. For each individual, we record the value for 6 disease events corresponding to 3 clinical measures, and we include 2 covariates. The motivation for including covariates is we are often interested in associations between different variables with disease severity. Each covariate was simulated to be positively correlated with disease stage. 

```{r TBebmData str}
data(TBebmData)
str(TBebmData)
```

### True event sequence
Here is the ground truth sequence of events that was used to generate TBebmData. There are 4 disease stages. Events X6, X1, and X3 occur simultaneously at stage 1: 

Event | Stage |
:-------:| :--------------: | 
X1 | 1 | 
X2 | 2 |
X3 | 1|
X4 | 3 | 
X5 | 4 | 
X6 | 1 | <br> 


**Important Note**      
  In theory we can include as many clinical measures and events as possible when fitting the EBM. The performance of the method decreases with smaller sample sizes, a larger number of events, and lower values of p for each event. We should consider these factors when choosing data to estimate the EBM. An additional challenge includes non-convergence when estimating the maximum likelihood sequence. This can potentially be mitigated by increasing the number of iterations during estimation but comes at the cost of increased computation time. 


### Data Summary 


Below we provide the first 6 rows of our example data set. The first observation in this data set is at disease stage 2. Events X1, X2, X3, and X6 have occurred as indicated with the value 1, and events X4 and X6 have not occurred which is indicated by the value 0. We generated this data set setting p to 0.9 so each individual’s data may not align perfectly with their true disease stage. For example, observation 3 and 4 are both at stage 1. Given the ground truth sequence, we would expect X1, X3, and X6 to all have the value 1. For observation 4 this is true, but for observation 3, X6 is equal to 0. 
```{r TBebmData first rows}
head(TBebmData)
```

Here is the distribution of disease stage. We can summarize true disease stage because we are using a simulated data set. We simulated disease stage originally using a uniform distribution, so we expect all stages to be approximately equally observed in the data. Notice here we have a disease stage 0 whereas in our true event sequence above we did not. This is because stage 0 is equivalent to no events having occurred. 
```{r stage dist}
prop.table(table(TBebmData$stage))
```

Here are the distributions for each covariate.
```{r data dist}
prop.table(table(TBebmData$C1))
fivenum(TBebmData$C2)
```
## Example analysis

### Visualize clinical events
Prior to fitting the model, it is good practice to first visualize the data we are using. We can plot our data using the plot_events function. This function first calculates the frequency of each event occurring and orders the events from most frequent to least frequent. The ordered events are the expected disease sequence for the data set as a greater event frequency should correspond to earlier disease stages. We then sort observations by the ordered events. We then plot each observation to visualize how well the data follow the expected disease sequence. The x-axis corresponds to observations after sorting and the y-axis corresponds to clinical events ordered by frequency. Blue indicates the event has occurred and white indicates the event has not occurred. If our data aligns well with the expected disease sequence, we expect the top left area of the plot to be mostly white and the bottom right to be consistently blue.  
  
We present the plot from  running the plot_events function with TBebmData. The y axis is ordered as X1, X6, X3, X2, X4, X5. This aligns with our ground truth sequence where X1, X6, and X3 occur at stage 1 followed by X2, X4, and X5 at stages 2, 3, and 4, respectively. We see our data align well with the ground truth sequence overall with few events occurring in the upper left hand portion of the plot. This is a result of us using a value of .9 for all clinical events when we simulated our data set. If p was lower, we would expect to see more variation. 
```{r plot events}
plot_events(TBebmData, id.var="Index", event.vars=c("X1", "X2", "X3", "X4", "X5", "X6"))
```

### Estimate model
We fit the EBM using the fit_ebm function. Prior to running the model, the data set should be formatted such that each event has a  recorded value of 1 to indicate the event has occurred or 0 otherwise. If there are any observations with missing data they are excluded prior to model fitting. Columns should be arranged such that the ordering within clinical measures is maintained. We must also specify a value of p for each event.
```{r measures}
measure_info <- data.frame(clinical_measure=c("measure1", "measure1", "measure2", "measure2", "measure2", "measure3"),event_number = c(1,2,1,2,3,1), event_name=c(c("X1", "X2", "X3", "X4", "X5", "X6")))

print(measure_info)
```
The fit_ebm function returns a data frame with 4 columns (measure, event_num, event_name, stage). The stage column is the maximum likelihood sequence. We present the maximum likelihood sequence for TBebmData. We ran the model with only 3 start points and 250 iterations but in practice the minimum number of start points should be 10 with at least 2000 iterations. From this output, the maximum likelihood sequence is ({X6, X1, X3}, {X2}, {X4},{X5}) which is the same as our ground truth sequence. 
```{r fit ebm}

#fit_ebm(TBebmData[,2:7], c(.9,.9,.9,.9,.9,.9), measure_info, 1, 3)

```

   
   
### Likelihood ascent

### Bootstrap resamples

### Positional variance diagram

### Predicted disease stage



## Conclusion     
In this tutorial, we illustrated how to use the TBebm package to analyze data to classify TB severity using clinical measures observed at time of TB diagnosis. With our example data set TBebmData, we demonstrated how to do the following:   
  
* Visualize clinical events   
* Estimate the disease event sequence   
* Check the likelihood ascent plot for model convergence  
* Generate bootstrap resamples and plot the PVD to visualize uncertainty  
* Calculate predicted disease stage and plot the distribution
     
       
We conclude with some considerations and best practices for practitioners wanting to analyze data with TBebm. 

* The method we present was motivated by prior work in event-based modeling that was largely used to understand the progression of neurodegenerative diseases. When applying our method to TB data, we are not concerned with drawing conclusions about disease progression. We are using the EBM to classify disease severity at time of diagnosis only. It would not be appropriate to comment on disease progression as we know many assumptions for the EBM are not met with TB. 
* Fitting the EBM requires collecting data that is strongly correlated with disease severity. For TB disease, we include measures that are well-established as predictors of treatment outcome and TB sterilization (cavitation, smear status, etc) and that are often used individually to stratify patients into categories that represent advanced versus early disease.   
* Fitting the EBM also requires the user to input a value for p for each event included in the model. With TB diagnostic measures and chest x-ray data, we expect p to be close to 1. The user should always run a sensitivity analysis where they decrease p from their original model fit to assess the sensitivity of the maximum likelihood sequence to p.   
* In theory we can include as many clinical measures and events as possible when fitting the EBM. The performance of the method decreases with smaller sample sizes, a larger number of events, and lower values of p for each event. We should consider these factors when choosing data to estimate the EBM. An additional challenge includes non-convergence when estimating the maximum likelihood sequence. This can potentially be mitigated by increasing the number of iterations during estimation but comes at the cost of increased computation time. The user should always check the likelihood ascent plot to determine if convergence was reached with the chosen number of iterations.   
* Finally, it is always best to visualize the observed data before fitting the model which also provides the user with an expected disease sequence to compare to the maximum likelihood estimate. This is especially helpful if we are using the EBM as a hypothesis generating tool and do not already have a strong understanding about disease sequence or severity.  





